\texttt{gwsnr}: A Python Package for Efficient SNR Calculation of GWs

### 2024-11-05 arxiv/paper.tex
mac

\fancyfoot[L]{\parbox[t]{0.98\headwidth}{\footnotesize{\sffamily Hemantakumar Phurailatpam, Otto Akseli Hannuksela, (2024). \texttt{gwsnr}:
A Python Package for Efficient SNR Calculation of Gravitational
Waves. \textit{PendingJournal of Open Source Software}, Pending(Pending), Pending. \url{https://doi.org/10.xxxxxx/draft}}}}

@@

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={gwsnr: A Python Package for Efficient SNR Calculation of Gravitational Waves},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\title{\texttt{gwsnr}: A Python Package for Efficient SNR Calculation of
Gravitational Waves}

@@

Gravitational waves, ripples in spacetime predicted by Einstein's theory
of General Relativity, have revolutionized astrophysics since their
first detection in 2015 (Abbott, B. P. et al. (2016a)). These waves are
emitted by cataclysmic events like the merging of binary black holes
(BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a unique
window into the cosmos. A critical aspect of gravitational wave analysis
is the Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a
gravitational wave signal relative to the background noise in a
detector, like LIGO (The LIGO Scientific Collaboration et al. (2015)) or
Virgo (Acernese et al. (2014)). This ratio is pivotal in confirming the
detection of gravitational waves and extracting astrophysical
information from them (Abbott, B. P. et al. (2016b)). However, specific
scenarios in gravitational wave research, particularly in simulations of
detectable gravitational wave events (Abbott et al. (2016)) and in
hierarchical Bayesian analysis (Thrane and Talbot (2019)) where
selection effects are considered, demand extensive and efficient
computation of SNR. This requirement presents a significant challenge,
as conventional computational approaches are typically time-consuming
and impractical for such specialized and large-scale analyses (Taylor
and Gerosa (2018), Davide, Pratten, and Vecchio (2020)).

@@

The \emph{\texttt{gwsnr}} Python package addresses the need for
efficient SNR computation in gravitational wave research. It
innovatively streamlines SNR calculations, enhancing accuracy and
efficiency with several advanced techniques. Firstly, it utilizes an
innovative interpolation method, employing a partial-scaling approach
for accurately interpolating the SNR of gravitational waves from
spin-less binary systems wrt to their mass parameters. Secondly, the
package features a noise-weighted inner product method, similar to that
in the \emph{\texttt{bilby}} package (Ashton, Gregory and Biscoveanu,
Sylvia and Cornish, Neil and Dal Canto, Isaac and Kumar, Prayush and
Meacher, Duncan and Middleton, Hannah and Mistry, Divyansh and Smith,
Rory and Stevenson, Tom (2022)), but enhanced with multiprocessing
capabilities. This integration allows for the parallel processing of
complex calculations, thereby expediting the SNR computation. Thirdly, a
trained Artificial Neural Network (ANN) model is incorporated for rapid
`probability of detection' (Pdet) estimation for BBH systems with spin
precession. Lastly, \emph{\texttt{gwsnr}} leverages the
\emph{\texttt{numba}}'s Just-In-Time (njit) compiler (Lam, Pitrou, and
Seibert (2022)), which optimizes performance by compiling Python code
into machine code at runtime, drastically reducing execution times.
Beyond these technical merits, \emph{\texttt{gwsnr}} stands out for its
user-friendly features and seamless integration with other related
software packages, making it not just a powerful tool but also an
accessible one for researchers. These attributes position
\emph{\texttt{gwsnr}} as an invaluable asset in gravitational wave data
analysis, particularly in simulations of detectable binary mergers,
calculation of merger rates, determining gravitational wave lensing
rates (Phurailatpam et al. (2024), Ng et al. (2024)), and in the
analysis of selection effects within hierarchical Bayesian frameworks.
The package thus represents a significant step forward in gravitational
wave research, enabling more precise and efficient exploration of the
universe through gravitational wave observations and simulations.
Additionally, \emph{\texttt{gwsnr}} is instrumental in the
\emph{\texttt{ler}} (Phurailatpam et al. (2024)) package for calculating
detectable rates of both lensed and unlensed gravitational wave events,
showcasing its utility in advanced gravitational wave studies.

@@

The \emph{\texttt{gwsnr}} package introduces the Partial Scaling method
for SNR calculations in spin-less binary systems. This method, rooted in
the FINDCHIRP algorithm (Allen et al. (2012)), focuses on non-spinning
IMR waveforms and particularly interpolates the Partial scaled SNR
(\(\rho_{1/2}\)) based on mass parameters (\(M_{tot},q\)).

@@

  \begin{itemize}
  \item
    For a simple inspiral waveform, the optimal SNR is given by,
    \[\rho = F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta, \psi) \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\]
  \item
    \(F\) is defined as a function of luminosity distance (\(D_l\)),
    chirp mass (\(\mathcal{M}\)), inclination angle (\(\iota\)),
    polarization angles (\(\psi\)), right ascension (\(\alpha\)), and
    declination (\(\delta\)). \(f\) is the frequency, \(f_{ISCO}\) is
    the last stable orbit frequency and \(S_n(f)\) is the detector's
    noise curve or power spectral density (psd).
  \item
    Then, partial scaled SNR:
    \(\rho_{1/2} = \sqrt{ 4\int_0^\infty \frac{f^{-7/3}}{S_n(f)}df } \approx \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\)
  \item
    For an spinless frequency-domain IMR (Inspiral-Merger-Ringdown)
    waveform with optimal SNR equal to \(\rho\):
    \(\rho_{1/2} = \rho\,/\, F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta, \psi)\)
  \item
    \(\rho_{1/2}\) is considered a function of \(M_{tot}\) and \(q\).
  \end{itemize}
\end{itemize}

\paragraph{Noise-Weighted Inner Product
Method}\label{noise-weighted-inner-product-method}

@@

These formulations highlight \emph{\texttt{gwsnr}}'s capability to
efficiently process diverse gravitational wave signals, enhancing data
analysis accuracy and efficiency.

@@

The \emph{\texttt{gwsnr}} package now incorporates an artificial neural
network (ANN) model, developed using \emph{\texttt{TensorFlow}} (Abadi
et al. (2015)) and \emph{\texttt{sklearn}} (Pedregosa et al. (2011)), to
rapidly estimate the `probability of detection' (Pdet) in binary black
hole (BBH) systems using the IMRPhenomXPHM waveform approximant. This
complex inspiral-merger-ringdown (IMR) waveform model accounts for
spin-precessing systems with subdominant harmonics. The ANN model is
especially useful when precise signal-to-noise ratio (SNR) calculations
are not critical, providing a quick and effective means of estimating
Pdet. This value indicates detectability under Gaussian noise by
determining if the SNR exceeds a certain threshold. Trained on a large
dataset from the \emph{\texttt{ler}} package, the ANN model uses
`partial scaled SNR' values as a primary input, reducing input
dimensionality from 15 to 5 and enhancing accuracy. This approach offers
a practical solution for assessing detectability under specified
conditions.

@@

\bibitem[\citeproctext]{ref-bilby}
Ashton, Gregory and Biscoveanu, Sylvia and Cornish, Neil and Dal Canto,
Isaac and Kumar, Prayush and Meacher, Duncan and Middleton, Hannah and
Mistry, Divyansh and Smith, Rory and Stevenson, Tom. 2022. {``{bilby: a
user-friendly Bayesian inference library}.''} \emph{GitHub Repository}.
GitHub. \url{https://github.com/GregoryAshton/Bilby}.

>>>>

\fancyfoot[L]{\parbox[t]{0.98\headwidth}{\footnotesize{\sffamily Hemantakumar Phurailatpam, Otto Akseli Hannuksela, (2024). \texttt{gwsnr}:
A Python Package for Efficient SNR Calculation of
GWs. \textit{PendingJournal of Open Source Software}, Pending(Pending), Pending. \url{https://doi.org/10.xxxxxx/draft}}}}

@@

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={gwsnr: A Python Package for Efficient SNR Calculation of GWs},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\title{\texttt{gwsnr}: A Python Package for Efficient SNR Calculation of
GWs}

@@

Gravitational waves (GWs), ripples in spacetime predicted by Einstein's
theory of General Relativity, have revolutionized astrophysics since
their first detection in 2015 (Abbott, B. P. et al. (2016a)). These
waves are emitted by cataclysmic events like the merging of binary black
holes (BBHs), binary neutron stars (BNSs) and BH-NS pairs, providing a
unique window into the cosmos. A critical aspect of GW analysis is the
Signal-to-Noise Ratio (SNR). SNR quantifies the strength of a GW signal
relative to the background noise in a detector, like LIGO (The LIGO
Scientific Collaboration et al. (2015)) or Virgo (Acernese et al.
(2014)). This ratio is pivotal in confirming the detection of GWs and
extracting astrophysical information from them (Abbott, B. P. et al.
(2016b)). However, specific scenarios in GW research, particularly in
simulations of detectable GW events (Abbott et al. (2016)) and in
hierarchical Bayesian analysis (Thrane and Talbot (2019)) where
selection effects are considered, demand extensive and efficient
computation of SNR. This requirement presents a significant challenge,
as conventional computational approaches, such as noise-weighted inner
product, are typically time-consuming and impractical for such
specialized and large-scale analyses (Taylor and Gerosa (2018), Davide,
Pratten, and Vecchio (2020)).

@@

The \emph{\texttt{gwsnr}} Python package addresses the need for
efficient SNR computation in GW research. It innovatively streamlines
SNR calculations, enhancing accuracy and efficiency with several
advanced techniques. Firstly, it utilizes an innovative interpolation
method, employing a partial-scaling approach for accurately
interpolating the SNR of GWs from spin-less binary systems wrt to their
mass parameters. Secondly, the package features a noise-weighted inner
product method, similar to that in the \emph{\texttt{bilby}} package
(Ashton, Gregory et al. (2022)), but enhanced with multiprocessing
capabilities. This integration allows for the parallel processing of
complex calculations, thereby expediting the SNR computation. Thirdly, a
trained Artificial Neural Network (ANN) model is incorporated for rapid
`probability of detection' (Pdet) estimation for BBH systems with spin
precession. Lastly, \emph{\texttt{gwsnr}} leverages the
\emph{\texttt{numba}}'s Just-In-Time (njit) compiler (Lam, Pitrou, and
Seibert (2022)), which optimizes performance by compiling Python code
into machine code at runtime, drastically reducing execution times.
Beyond these technical merits, \emph{\texttt{gwsnr}} stands out for its
user-friendly features and seamless integration with other related
software packages, making it not just a powerful tool but also an
accessible one for researchers. These attributes position
\emph{\texttt{gwsnr}} as an invaluable asset in GW data analysis,
particularly in simulations of detectable compact binary mergers,
determining rates of GW events, ascertain to both lensed and un-lensed
events (Phurailatpam et al. (2024), Ng et al. (2024)), and in the
analysis of selection effects within hierarchical Bayesian frameworks
(Thrane and Talbot (2019)). The package thus represents a significant
step forward in aiding GW research, enabling more precise and efficient
exploration of the universe through GW observations and simulations.
Additionally, \emph{\texttt{gwsnr}} is instrumental in the
\emph{\texttt{ler}} (Phurailatpam et al. (2024)) package for calculating
detectable rates of both lensed and unlensed GW events, showcasing its
utility in advanced GW studies.

@@

The \emph{\texttt{gwsnr}} package introduces the Partial Scaling method
for SNR calculations in spin-less binary systems. This method, rooted in
the FINDCHIRP algorithm (Allen et al. (2012)), focuses on non-spinning
inspiral-merger-ringdown (IMR) waveforms, in lalsimulation library (LIGO
Scientific Collaboration, Virgo Collaboration, and KAGRA Collaboration
(2018)), and particularly interpolates the Partial scaled SNR
(\(\rho_{1/2}\)) based on mass parameters (\(M_{tot},q\)).

@@

  \begin{itemize}
  \item
    For a simple inspiral waveform, the optimal SNR is given by,
    \[\rho = F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta) \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\]
  \item
    \(F\) is defined as a function of luminosity distance (\(D_l\)),
    chirp mass (\(\mathcal{M}\)), inclination angle (\(\iota\)),
    polarization angles (\(\psi\)), right ascension (\(\alpha\)), and
    declination (\(\delta\)). \(f\) is the frequency, \(f_{ISCO}\) is
    the last stable orbit frequency and \(S_n(f)\) is the detector's
    noise curve or power spectral density (psd).
  \item
    Then, partial scaled SNR:
    \(\rho_{1/2} = \sqrt{ 4\int_0^\infty \frac{f^{-7/3}}{S_n(f)}df } \approx \sqrt{ 4\int_0^{f_{ISCO}} \frac{f^{-7/3}}{S_n(f)}df }\)
  \item
    For an spinless frequency-domain IMR waveform with optimal SNR equal
    to \(\rho\):
    \(\rho_{1/2} = \rho\,/\, F(D_l,\mathcal{M},\iota,\psi,\alpha, \delta, \psi)\)
  \item
    \(\rho_{1/2}\) is considered a function of \(M_{tot}\) and \(q\).
  \end{itemize}
\end{itemize}

\paragraph{Noise-Weighted Inner Product Method with
Multiprocessing}\label{noise-weighted-inner-product-method-with-multiprocessing}

@@

These formulations highlight \emph{\texttt{gwsnr}}'s capability to
efficiently process diverse GW signals, enhancing data analysis accuracy
and efficiency.

@@

The \emph{\texttt{gwsnr}} package now incorporates an artificial neural
network (ANN) model, developed using \emph{\texttt{TensorFlow}} (Abadi
et al. (2015)) and \emph{\texttt{sklearn}} (Pedregosa et al. (2011)), to
rapidly estimate the Pdet in binary black hole (BBH) systems using the
IMRPhenomXPHM waveform approximant. This complex IMR waveform model
accounts for spin-precessing systems with subdominant harmonics. The ANN
model is especially useful when precise signal-to-noise ratio (SNR)
calculations are not critical, providing a quick and effective means of
estimating Pdet. This value indicates detectability under Gaussian noise
by determining if the SNR exceeds a certain threshold. Trained on a
large dataset from the \emph{\texttt{ler}} package, the ANN model uses
`partial scaled SNR' values as a primary input, reducing input
dimensionality from 15 to 5 and enhancing accuracy. This approach offers
a practical solution for assessing detectability under specified
conditions.

@@

\bibitem[\citeproctext]{ref-bilby}
Ashton, Gregory, Sylvia Biscoveanu, Neil Cornish, Isaac Dal Canto,
Prayush Kumar, Duncan Meacher, Hannah Middleton, Divyansh Mistry, Rory
Smith, and Tom Stevenson. 2022. {``{bilby: a user-friendly Bayesian
inference library}.''} \emph{GitHub Repository}. GitHub.
\url{https://github.com/GregoryAshton/Bilby}.
